---
title: "Deconvolution in 1D"
format: gfm
editor: source
---

This vignette walks through deconvolution in 1D.

```{r}
library(data.table)
library(dplyr)
library(tidyr)
library(ggplot2)
library(dcon)
set.seed(13579)
```

# Quick start

We first provide a quick example to show how the method works. For more
details, see everything below the quick start section.

Here, we use a small sample of the
RD-SPRITE data 
([GSE151515](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE151515)).

We focus on the lncRNA \textit{Airn} which is a part of the \textit{Igf2r}
imprinted cluster on chromosome 17 (mm10) and is thought to orchestrate 
silencing of the gene  \textit{Slc22a3}.


```{r load-data}
airn <- fread('/rafalab/lzou/rdsprite/count_windows/Airn_RNA_count_windows_10000_clusters_2-1000_chr17_129S1_SvImJ.csv') |>
  filter(chrom=='chr17')
```

```{r, fig.height=3, fig.width=8}
airn |>
  filter(start <= 30e6) |>
  ggplot(aes(x = start/1e6, y = count)) +
  geom_rect(xmin = 12739311/1e6, xmax = 12861884/1e6,
            ymin = 0, ymax = max(airn$count),
            fill = 'lightpink') +
  geom_rect(xmin = 12417972/1e6, xmax = 12509704/1e6,
            ymin = 0, ymax = max(airn$count),
            fill = 'lightgreen') +
  geom_point(size=0.1) +
  scale_y_continuous(expand=c(0,0)) +
  scale_x_continuous(expand=c(0,0)) +
  theme_classic() +
  xlab('position along chr17 (mb)') +
  ylab('Airn lncRNA raw count')
```

# Model

The model for deconvolution is

$$Y \sim \text{Poisson}(D\lambda)$$
$$\text{log}(\lambda) = B\alpha$$

where $D$ represents the DNA-DNA contact matrix and $\lambda$ the true,
underlying signal. $B$ are Gaussian basis functions and $\alpha$ are the
coefficients to be estimated.

# Preprocessing: DNA-DNA matrix

Before we deconvolve, it is important to inspect the DNA-DNA contact matrix and
preprocess it.

Here, we use an example DNA matrix, taken from the
\textit{Kcnq1ot1} imprinted region (a 1mb region around the \textit{Kcnq1ot1}
locus) at 10kb bin resolution.

```{r, fig.height=4, fig.width=5}
dd <- readRDS(system.file("extdata", "rdsprite_kcnq1ot1_domain_DD.rds", package="dcon"))
dd <- as.matrix(dd)
plot_mat(dd, main = expression(paste('Raw DNA-DNA contacts at the ', italic('Kcnq1ot1'), ' locus (1mb)')))
```


## Normalization and smoothing

There are a few important tuning parameters to normalize and smooth the 
DNA-DNA contact matrix with `normalize_hic`.

1. `threshold` -  for setting a threshold for the total number of
contacts necessary per column to be considered for deconvolution; otherwise it
can be quite noisy if there are gaps. For example,
note that the center cross where there is no signal is because of the repetitive
LINE element located inside of the Kcnq1ot1 locus, which is difficult to map
reads to. The default threshold is set to be the bottom 0.5% quantile of the
`colSums` of the matrix, but this can also be manually adjusted with `threshold=`.
2. `smooth` - `TRUE` or `FALSE`, this uses a Gaussian blur to smooth the matrix.
This takes care of instances like above where such mapping discrepancies can 
result in 'holes' in the matrix.
3. `sigma` - the SD of the Gaussian blur which controls the amount of smoothing 
to apply. Default is `sigma=1`.

Finally, after thresholding and smoothing, the matrix is column-normalized. 

We illustrate the importance of these steps below:

With no thresholding, we see that there are indeed some spurious reads in those
'empty' gaps, and if we just column-normalize, we end up getting a messy
signal:

```{r, fig.height=4, fig.width=5}
ddnorm <- normalize_hic(dd, threshold=0)
plot_mat(ddnorm, main = 'Column-normalized, with no thresholding or smoothing')
```


The default amount of thresholding clears out the signal in the lowest 0.5%
of bins and sets all the signal to be on the diagonal. Smoothing helps essentially 
impute nearby values that are lost. Note that we choose not to make any
assumptions about how the 'empty' region interacts with other regions; 
effectively, it only spreads locally to nearby bins.

```{r, fig.height=4, fig.width=5}
ddnorm <- normalize_hic(dd, smooth = T)
plot_mat(ddnorm, main = 'Column-normalized, with thresholding and smoothing')
```

## Setting gamma, the amount of deconvolution

In practice, even after thresholding and smoothing, the DNA-DNA contact matrix
can still be quite noisy. The `gamma` parameter is important to set because it
controls a penalty on the off-diagonal of the DNA-DNA contact matrix, 
essentially controlling how much of the original signal we want to preserve
vs. deconvolve. We have $$D=(1-\gamma)I+\gamma O,$$, where $O$ is the smoothed,
column-normalized matrix we created above. We will explore the effects of 
different values of `gamma` below.


# Simulated data

## Gamma and degrees of freedom 

We now generate a random true and observed signal assuming different `gamma` values,
and assuming we use different `gamma` values to deconvolve.

In general, using a lower degrees of freedom, set using `df`, results in fewer
spurious peaks while still capturing the main true signal.


```{r}
par(mfrow=c(2,2))
for (g in c(0.25, 0.5, 0.75, 0.9)) {
  ddnorm <- normalize_hic(dd, gamma = g, smooth = T)
  sim <- dcon:::simulate_y(len = 100, df = 10, D = ddnorm, npeaks=1)
  fit0.25 <- fit_decon(sim$y, normalize_hic(dd, gamma = 0.25, smooth = T), df = 10)
  fit0.5 <- fit_decon(sim$y, normalize_hic(dd, gamma = 0.5, smooth = T), df = 10)
  fit0.75 <- fit_decon(sim$y, normalize_hic(dd, gamma = 0.75, smooth = T), df = 10)
  fit0.9 <- fit_decon(sim$y, normalize_hic(dd, gamma = 0.9, smooth = T), df = 10,
                      hessian = T)
  plot(1:100, sim$y, ylim=c(0,200), main = paste0('gamma=',g))
  lines(1:100, exp(sim$B%*%sim$a), col='red')
  lines(1:100, exp(fit0.25$est), col='#8cb0fb')
  lines(1:100, exp(fit0.5$est), col='#748cd0')
  lines(1:100, exp(fit0.75$est), col='#5b6aa6')
  lines(1:100, exp(fit0.9$est), col='#424b7e')
}
```

```{r}
for (g in c(0.25, 0.5, 0.75, 0.9)) {
  ddnorm <- normalize_hic(dd, gamma = g, smooth = T)
  sim <- dcon:::simulate_y(len = 100, df = 10, D = ddnorm, npeaks=1)
  fit0.25 <- fit_decon(sim$y, normalize_hic(dd, gamma = 0.25, smooth = T), df = 5)
  fit0.5 <- fit_decon(sim$y, normalize_hic(dd, gamma = 0.5, smooth = T), df = 5)
  fit0.75 <- fit_decon(sim$y, normalize_hic(dd, gamma = 0.75, smooth = T), df = 5)
  fit0.9 <- fit_decon(sim$y, normalize_hic(dd, gamma = 0.9, smooth = T), df = 5,
                      hessian = T)
  plot(1:100, sim$y, ylim=c(0,200), main = paste0('gamma=',g))
  lines(1:100, exp(sim$B%*%sim$a), col='red')
  lines(1:100, exp(fit0.25$est), col='#8cb0fb')
  lines(1:100, exp(fit0.5$est), col='#748cd0')
  lines(1:100, exp(fit0.75$est), col='#5b6aa6')
  lines(1:100, exp(fit0.9$est), col='#424b7e')
}
```






```{r}
sessionInfo()
```




